---
title: "Atlas test"
author: "Wojciech Banas"
date: "2025-02-06"
output: pdf_document
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r, include=F}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# Package load

```{r}
#install.packages("pacman")
pacman::p_load(dplyr, kableExtra, 
               ggplot2, stringr,
               ggsci, data.table, ggmap,
               lubridate, rnaturalearthdata,
               sf, rnaturalearth, broom,
               readxl)

options(scipen = 9) # disable scientific notation
`%nin%` <- purrr::negate(`%in%`)
theme_set(theme_bw())

```

# Data load 

```{r}
# Set the directory containing the files- assumes the unpacked test_data is in the same folder as this .rmd file
directory_path <- "test_data/data/dest"

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
# Get a list of all files in the directory
file_list <- list.files(directory_path, full.names = TRUE)

# Function create clean names for files
clean_name <- function(filename) {
  name <- basename(filename)
  name <- str_remove(name, "\\.csv\\.gz$")  # Remove .csv.gz
  name <- str_remove(name, "\\.csv$")  # Remove .csv
  name
}

# Loop through each file and load it into memory
for (file in file_list) {
  file_name <- clean_name(file)  # Get cleaned file name
  cat("Loading file: \n", file_name, "\n")
  if (str_detect(file, "\\.csv\\.gz$")) {
    assign(file_name, read.csv(gzfile(file)), envir = .GlobalEnv)  # Read .csv.gz
  } else if (str_detect(file, "\\.csv$")) {
    assign(file_name, read.csv(file), envir = .GlobalEnv)  # Read .csv
  }
}


# Read UK population estimates from https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland


pop_estimates <- read_xlsx("ukpopulationestimates183820231.xlsx",sheet = 7, skip = 1)

```

# Task 1 

Instructions:
The data files contain errors and bugs that reflect common problems and challenges associated with electronic health records. Load and clean the data, both individually and linked. Where patients fail a quality control check, exclude them from the analyses. The output of this task is a clean clean patient demographics table for use in the next steps and sanitized condition, observation and medication files.

Goals:
*   Remove patients with erroneous birth or death dates from the patients table or sex (not always essential to remove, but here I assume it is, given that we need to calculate adjusted prevalence). Missing data from other fields in that table is left as is, as these aren't explicitly relevant. 
*   Remove entries from other tables that do not map to the patients table 
*   Remove duplicated entries
*   Remove entries that do not match the expected field type 


## Sanity checks and cleanup 

Here I go through each table, have a look how many patients it describes, and doing some pre-processing if needed.

### Patients 

*   Removing patients with incorrect birth/death dates
*   Removing patients with incorrect gender data 
*   Cleaning up ethnicity 
*   Providing overview of N unique patients

```{r}
#str(patients)

cat("N rows in patients table:")
nrow(patients)

cat("N unique patients in patients table:")
uniqueN(patients$Id)

cat("N unique entries by ID and passport:")
nrow(distinct(patients, 
              Id, PASSPORT))

cat("No duplicated patient IDs detected")


cat("Errors in birth/death dates:")

patients %>%
  mutate(across(c(BIRTHDATE, DEATHDATE), ~ ymd(.x), .names = "DATE_{.col}"),
         to_remove = case_when(is.na(DATE_BIRTHDATE) ~ "Erroneous birth date",
                               is.na(DATE_DEATHDATE) & (DEATHDATE != "") ~ "Erroneous death date",
                               DATE_BIRTHDATE > ymd("2025-01-01") ~ "Born in 2025 or later",# Assuming birth dates in 2025 or later are erroneous
                               DATE_BIRTHDATE < ymd("1850-01-01") ~ "Born before 1850",# Assuming birth dates before 1850 are erroneous
                               DATE_BIRTHDATE > DATE_DEATHDATE ~ "Died before birth",
                               T~ "No errors")) %>%
  pull(to_remove) %>%
  table()

cat("These records are removed- assuming correct date of birth is crucial for this project\n given we need to calculated adjusted prevalence")

patients_clean <- patients %>%
  mutate(across(c(BIRTHDATE, DEATHDATE), ~ ymd(.x), .names = "DATE_{.col}"),
         to_remove = case_when(is.na(DATE_BIRTHDATE) ~ "Erroneous birth date",
                               is.na(DATE_DEATHDATE) & (DEATHDATE != "") ~ "Erroneous death date",
                               DATE_BIRTHDATE > ymd("2025-01-01") ~ "Born in 2025 or later",# Assuming birth dates in 2025 or later are erroneous
                               DATE_BIRTHDATE < ymd("1850-01-01") ~ "Born before 1850",# Assuming birth dates before 1850 are erroneous
                               DATE_BIRTHDATE > DATE_DEATHDATE ~ "Died before birth",
                               T~ "No errors")) %>%
  filter(to_remove == "No errors") %>%
  select(-to_remove, -BIRTHDATE, -DEATHDATE)

cat("Unique levels of race:")
patients_clean %>%
  pull(RACE) %>%
  table()

cat("Levels ``???-`` and ``XJniDSe`` are relabelled to ``unknown``")
patients_clean <- patients_clean %>%
  mutate(RACE = ifelse(RACE %in% c("???-",
                                   "XJniDSe"),
                       "Unknown",
                       RACE))


cat("Unique levels of gender:")
patients_clean %>%
  pull(GENDER) %>%
  table()

cat("Levels ``8293.3 `` and `` `` are  removed. \n Assuming correct gender is crucial as we need to calculate adjusted prevalence")
patients_clean <- patients_clean %>%
  mutate(GENDER = ifelse(GENDER %in% c("8293.3",
                                   ""),
                       "Unknown",
                       GENDER)) %>%
  filter(GENDER != "Unknown")

cat("N unique patients in the new patients table:")
uniqueN(patients_clean$Id)
```

# conditions 

*   Providing overview of N rows and patients 
*   Removing records that do not match patients in patients table


```{r}
cat("N rows of data in conditions table")
nrow(conditions)
cat("N rows of distinct data in conditions table")
nrow(distinct(conditions))

cat("N duplicated rows removed:")
nrow(conditions) - nrow(distinct(conditions))

conditions_clean <- distinct(conditions)


cat("N records in conditions table without a match in patients table:")
sum(unique(conditions_clean$PATIENT) %nin% unique(patients_clean$Id))

cat("Records of that patient are removed")
patient_to_remove <- unique(conditions$PATIENT)[unique(conditions_clean$PATIENT) %nin% unique(patients_clean$Id)]
conditions_clean <- conditions_clean %>%
  filter(PATIENT %nin% patient_to_remove)

cat("N records with incorrect date")
conditions_clean <- conditions_clean %>%
  mutate(across(c(START, STOP), ~ as.Date(.)))

conditions_clean %>%
  filter(is.na(START)) %>%
  nrow()
conditions_clean %>%
  filter(is.na(STOP)) %>%
  nrow()

cat("N records in patients table without a match in conditions table:")
sum(unique(patients_clean$Id) %nin% unique(conditions_clean$PATIENT))
```


# observations

*   Providing overview of N rows and patients 
*   Removing records that do not match patients in patients table

```{r}
cat("N rows of data in observations table")
nrow(observations)
cat("N rows of distinct data in observations table")
nrow(distinct(observations))

cat("N records in observations table without a match in patients table:")
sum(unique(observations$PATIENT) %nin% unique(patients_clean$Id))

cat("Records of that patient are removed")
patient_to_remove <- unique(observations$PATIENT)[unique(observations$PATIENT) %nin% unique(patients_clean$Id)]
observations_clean <- observations %>%
  filter(PATIENT %nin% patient_to_remove)

cat("N rows without a linked encounter")

cat("N records in patients table without a match in observations table:")
sum(unique(patients_clean$Id) %nin% unique(observations_clean$PATIENT))

observations_clean <- observations_clean %>%
  mutate(DATE = ymd_hms(DATE)) 

cat("N records with incorrect date")
observations_clean %>%
  filter(is.na(DATE)) %>%
  nrow()

cat("N records without a valid encounter ID, which are removed")
observations_clean %>%
  filter(ENCOUNTER == "") %>% nrow()

observations_clean <- observations_clean %>%
  filter(ENCOUNTER != "")
```

# medications

*   Providing overview of N rows and patients 
*   Removing records that do not match patients in patients table

```{r}
cat("N rows of data in medications table")
nrow(medications)
cat("N rows of distinct data in medications table")
nrow(distinct(medications))

cat("N records in observations table without a match in patients table:")
sum(unique(medications$PATIENT) %nin% unique(patients_clean$Id))

cat("Records of that patient are removed")
patient_to_remove <- unique(medications$PATIENT)[unique(medications$PATIENT) %nin% unique(patients_clean$Id)]
medications_clean <- medications %>%
  filter(PATIENT %nin% patient_to_remove)

medications_clean <- medications_clean %>%
  mutate(START = ymd_hms(START)) 

cat("N records with incorrect date")
medications_clean %>%
  filter(is.na(START)) %>%
  nrow()


cat("N records in patients table without a match in observations table:")
sum(unique(patients_clean$Id) %nin% unique(medications_clean$PATIENT))
```



# Task 2 

 Provide a comprehensive description of the patients, observations, and conditions tables - for example, provide the number of unique patients, the most frequent ontology terms (e.g. LOINC, SNOMED and RxNorm) and other information you think is important to describe.

## Demographics table 

*   Creating a table describing ethnicity, race, gender, age, death
*   Plotting patients on a map

```{r}
n_patients <- nrow(patients_clean)

pat_ethn <- patients_clean %>% 
  group_by(ETHNICITY) %>%
  summarise(n = n()) %>%
  rename(item = ETHNICITY)

pat_race <- patients_clean %>% 
  group_by(RACE) %>%
  summarise(n = n())%>%
  rename(item = RACE)

pat_gend <- patients_clean %>%
  group_by(GENDER) %>%
  summarise(n = n())%>%
  rename(item = GENDER)

pat_died <- patients_clean %>%
  mutate(dead = !is.na(DATE_DEATHDATE)) %>%
  group_by(dead) %>%
  summarise(n = n())%>%
  rename(item = dead)

#observations_clean$DATE %>% max()
#encounters$START %>% ymd_hms() %>% max()

patients_clean <- patients_clean %>%
  mutate(end_date = as.Date("2021-11-19"), # assuming this is the extract end date, as the max dates of observations and encounters is then
         age = ifelse(!is.na(DATE_DEATHDATE),
                      difftime(DATE_DEATHDATE, DATE_BIRTHDATE),
                      difftime(end_date, DATE_BIRTHDATE)),
         age = age/365.25,
         age_group = cut(age, 
                         breaks = c(seq(0, 90, by = 10), Inf), 
                         labels = c("0-9", "10-19", "20-29", "30-39", "40-49", 
                                    "50-59", "60-69", "70-79", "80-89", "90 or more"),
                         right = FALSE),
         age_group = factor(age_group, levels = c("0-9", "10-19", "20-29", "30-39", "40-49", 
                                    "50-59", "60-69", "70-79", "80-89", "90 or more"), 
                            ordered = TRUE))

pat_age <- patients_clean %>%
  group_by(age_group) %>%
  summarise(n = n()) %>%
  rename(item = age_group)

rbind(pat_ethn, pat_race, pat_gend, pat_died, pat_age) %>%
  mutate(perc = n/n_patients) %>%
  rowwise() %>%
  mutate(perc = paste0(n, " (", round(100*perc,2), ")")) %>%
  ungroup() %>%
  rename(`N (%)` = perc) %>%
  select(-n) %>%
  kbl(booktabs = T,
      caption = paste0("Demographics table of the population (N = ", n_patients, ")")) %>%
  pack_rows("Ethnicity", 1, 
            nrow(pat_ethn)) %>%
  pack_rows("Race", nrow(pat_ethn) +1,
            nrow(pat_ethn)  +nrow(pat_race)) %>%
  pack_rows("Gender", 1+nrow(pat_ethn)+nrow(pat_race),
            nrow(pat_ethn)+nrow(pat_race)+ nrow(pat_gend)) %>%
  pack_rows("Died", 1+nrow(pat_ethn)+nrow(pat_race)+ nrow(pat_gend),
            nrow(pat_ethn)+nrow(pat_race)+ nrow(pat_gend)+nrow(pat_died))%>%
  pack_rows("Age", 1+ nrow(pat_ethn)+nrow(pat_race)+ nrow(pat_gend)+nrow(pat_died),
            nrow(pat_ethn)+nrow(pat_race)+ nrow(pat_gend)+nrow(pat_died)+nrow(pat_age))




# Load a map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Convert data to spatial format
patients_sf <- st_as_sf(patients_clean, coords = c("LON", "LAT"), crs = 4326)

# Plot the map
ggplot() +
  geom_sf(data = world) +
  geom_sf(data = patients_sf, color = "red", size = 3) +
  labs(title = "Patient Locations", x = "Longitude", y = "Latitude")

bbox <- st_bbox(patients_sf) 
ggplot() +
  geom_sf(data = world) +
  geom_sf(data = patients_sf, color = "red", size = 3) +
    coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]), ylim = c(bbox["ymin"], bbox["ymax"]), expand = TRUE) +
  labs(title = "Patient Locations", x = "Longitude", y = "Latitude")
```

## conditions

*   Cleaning up SNOMED dictionary and joining it to the conditions table 
*   Providing an overview of the conditions 
*   Histogram of the number of conditions (observations) per patient
*   Histogram of the events over time


```{r}
# first, clean up the snomed dictionary 
duped_snomed_codes <- dictionary_snomed$CODE[dictionary_snomed$CODE %>%
  duplicated()]
to_remove <- dictionary_snomed %>%
  filter(CODE %in% duped_snomed_codes) %>%
  group_by(CODE) %>%
  mutate(n_th = row_number()) %>%
  filter(n_th >1) # assuming no difference between e.g. "Epilepsy (disorder)" and "Epilepsy"

dictionary_snomed_clean <- dictionary_snomed %>%
  filter(DESCRIPTION %nin% to_remove$DESCRIPTION)

conditions_clean <- conditions_clean %>%
  left_join(dictionary_snomed_clean, by = "CODE")

patients_per_condition <- conditions_clean %>%
  distinct(PATIENT, DESCRIPTION) %>%
  pull(DESCRIPTION) %>%
  table() %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>% 
  mutate(perc = Freq/uniqueN(patients_clean$Id)) %>%
  rowwise() %>%
  mutate(Freq_perc = paste0(Freq, " (", round(perc*100, 2), ")")) %>%
  ungroup() %>%
  select(-perc) %>%
  rename("SNOMED item" = ".",
         "N patients (%)" = Freq_perc)

head(patients_per_condition, 10) %>%
  select(-Freq) %>%
  kbl(booktabs = T,
      caption = paste0("Prevalence of the top 10 most frequent SNOMED items" , "(N = ", n_patients, ")"))

patients_per_condition %>%
  filter(Freq == min(Freq)) %>%
  select(-Freq) %>%
  kbl(booktabs = T,
      caption = paste0("Prevalence of the least frequent SNOMED items" , " (N = ", n_patients, ")"))

conditions_clean %>%
  distinct(PATIENT, DESCRIPTION) %>%
  group_by(PATIENT) %>%
  mutate(n = n()) %>%
  distinct(PATIENT, n) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = (conditions_clean %>%
                           distinct(PATIENT, DESCRIPTION) %>%
                           group_by(PATIENT) %>%
                           mutate(n = n()) %>%
                           distinct(PATIENT, n) %>% pull(n) %>% max())) +
  labs(title = "Histogram of the number of unique SNOMED items per patient", 
       x = "Number of unique diagnoses")

conditions_clean %>%
  ggplot(aes(START)) +
  geom_histogram(bins = 300) +
  labs(title = paste0("Histogram of the timing of events in the conditions dataset. \n Min: ",
                      min(conditions_clean$START), ", max: ", max(conditions_clean$START)))

```

## observations

*   Cleaning up loinc dictionary and joining it to the observations table 
*   Providing an overview of the observations 
*   Histogram of the number of observations per patient
*   Histogram of the events over time

```{r}
cat("Categories of observations")
observations_clean$CATEGORY %>%
  table()

cat("Types of observations")
observations_clean$TYPE %>%
  table()

cat("Types of Units")
observations_clean$UNITS %>%
  table()

cat("Number of unique text-type observations")
observations_clean %>% 
  filter(TYPE == "text") %>%
  pull(VALUE) %>%
  uniqueN()

# Numeric observations convert to numeric without errors
#observations_clean %>%
 # filter(TYPE == "numeric") %>%
  #mutate(VALUE = as.numeric(VALUE))

duped_loinc_codes <- dictionary_loinc$CODE[dictionary_loinc$CODE %>%
  duplicated()]
to_remove <- dictionary_loinc %>%
  filter(CODE %in% duped_loinc_codes) %>%
  arrange(CODE) %>%
  group_by(CODE) %>%
  mutate(string_length = str_length(DESCRIPTION)) %>%
  filter(string_length != max(string_length)) # assuming the correct description is the longer one 

dictionary_loinc_clean <- dictionary_loinc %>%
  filter(DESCRIPTION %nin% to_remove$DESCRIPTION) 

observations_clean <- observations_clean %>%
  left_join(dictionary_loinc_clean, by = "CODE")


observations_per_patient <- observations_clean %>%
  distinct(PATIENT, DESCRIPTION) %>%
  pull(DESCRIPTION) %>%
  table() %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>% 
  mutate(perc = Freq/uniqueN(patients_clean$Id)) %>%
  rowwise() %>%
  mutate(Freq_perc = paste0(Freq, " (", round(perc*100, 2), ")")) %>%
  ungroup() %>%
  select(-perc) %>%
  rename("loinc item" = ".",
         "N patients (%)" = Freq_perc)


head(observations_per_patient, 10) %>%
  select(-Freq) %>%
  kbl(booktabs = T,
      caption = paste0("Prevalence of the top 10 most frequent loinc items" , "(N = ", n_patients, ")"))

observations_per_patient %>%
  filter(Freq == min(Freq)) %>%
  select(-Freq) %>%
  kbl(booktabs = T,
      caption = paste0("Prevalence of the least frequent loinc items" , " (N = ", n_patients, ")"))

observations_clean %>%
  distinct(PATIENT, DESCRIPTION) %>%
  group_by(PATIENT) %>%
  mutate(n = n()) %>%
  distinct(PATIENT, n) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = (observations_clean %>%
                           distinct(PATIENT, DESCRIPTION) %>%
                           group_by(PATIENT) %>%
                           mutate(n = n()) %>%
                           distinct(PATIENT, n) %>% pull(n) %>% max())) +
  labs(title = "Histogram of the number of unique loinc items per patient", 
       x = "Number of unique diagnoses")


observations_clean %>%
  ggplot(aes(DATE)) +
  geom_histogram(bins = 300) +
  labs(title = paste0("Histogram of the timing of events in the observations dataset. \n Min: ",
                      min(observations_clean$DATE), ", max: ", max(observations_clean$DATE)))



```


## Medications

*   Cleaning up rxnorm dictionary and joining it to the observations table 
*   Providing an overview of the medications 
*   Histogram of the number of medications per patient
*   Histogram of the events over time

```{r}
# first, clean up the rxnorm dictionary 
duped_rxnorm_codes <- dictionary_rxnorm$CODE[dictionary_rxnorm$CODE %>%
  duplicated()]

to_remove <- dictionary_rxnorm %>%
  filter(CODE %in% duped_rxnorm_codes) %>%
  group_by(CODE) %>%
  mutate(n_th = row_number()) %>%
  filter(n_th >1) # assuming no difference between the duplicates

dictionary_rxnorm_clean <- dictionary_rxnorm %>%
  mutate(CODE = as.character(CODE)) %>%
  filter(DESCRIPTION %nin% to_remove$DESCRIPTION)

medications_clean <- medications_clean %>%
  mutate(CODE = str_remove(CODE, "\\.0$")) %>%
  left_join(dictionary_rxnorm_clean, by = "CODE")

cat("N observations in medications that have no valid description, which will be removed")
medications_clean %>% 
  filter(is.na(DESCRIPTION)) %>% nrow()

medications_clean <- medications_clean %>%
  filter(!is.na(DESCRIPTION))

medication_per_patient <- medications_clean %>%
  distinct(PATIENT, DESCRIPTION) %>%
  pull(DESCRIPTION) %>%
  table() %>%
  as.data.frame() %>%
  arrange(desc(Freq)) %>% 
  mutate(perc = Freq/uniqueN(patients_clean$Id)) %>%
  rowwise() %>%
  mutate(Freq_perc = paste0(Freq, " (", round(perc*100, 2), ")")) %>%
  ungroup() %>%
  select(-perc) %>%
  rename("SNOMED item" = ".",
         "N patients (%)" = Freq_perc)

head(medication_per_patient, 10) %>%
  select(-Freq) %>%
  kbl(booktabs = T,
      caption = paste0("Prevalence of the top 10 most frequent rxnorm items" , "(N = ", n_patients, ")")) %>%
  kable_styling(latex_options = "scale_down")

medication_per_patient %>%
  filter(Freq == min(Freq)) %>%
  select(-Freq) %>%
  kbl(booktabs = T,
      caption = paste0("Prevalence of the least frequent rxnorm items" , " (N = ", n_patients, ")"))

medications_clean %>%
  distinct(PATIENT, DESCRIPTION) %>%
  group_by(PATIENT) %>%
  mutate(n = n()) %>%
  distinct(PATIENT, n) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = (medications_clean %>%
                           distinct(PATIENT, DESCRIPTION) %>%
                           group_by(PATIENT) %>%
                           mutate(n = n()) %>%
                           distinct(PATIENT, n) %>% pull(n) %>% max())) +
  labs(title = "Histogram of the number of unique rxnorm items per patient", 
       x = "Number of unique diagnoses")



medications_clean %>%
  ggplot(aes(START)) +
  geom_histogram(bins = 300) +
  labs(title = paste0("Histogram of the timing of events in the medications dataset. \n Min: ",
                      min(medications_clean$START), ", max: ", max(medications_clean$START)))


```



# Task 3a

Using the cleaned data from the first task, explore and compare the distribution of systolic and diastolic blood pressure

*   Providing the number of observations
*   Showing the histogram and the potentially erroneous data, which are then removed 
*   Showing updated histogram and overlaid density
*   Showing correlation between systolic and diastolic BP per encounter
*   Showing summary statistics for BP


```{r}
cat("N observations of either systolic or diastolic BP")

task_3a <- observations_clean %>%
  filter(str_detect(DESCRIPTION, "Diastolic|Systolic")) %>%
  mutate(VALUE = as.numeric(VALUE))

task_3a %>%
  pull(DESCRIPTION) %>%
  table()


task_3a %>%
  ggplot(aes(VALUE)) +
  geom_histogram(bins = 300) +
  facet_wrap(~DESCRIPTION)

cat("N unique observations of Diastolic BP <50 and its values")
task_3a %>%
  filter(DESCRIPTION == "Diastolic Blood Pressure") %>%
  filter(VALUE < 50) %>%
  group_by( VALUE) %>%
  summarise(n = n()) %>%
  arrange(desc(n))


cat("N unique observations of Systolic BP > 270 and its values")

task_3a %>%
  filter(DESCRIPTION == "Systolic Blood Pressure") %>%
  filter(VALUE > 270) %>%
  group_by( VALUE) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

cat("N observations of Systolic BP equal 300. These observations are removed")

task_3a %>%
  filter(DESCRIPTION == "Systolic Blood Pressure") %>%
  filter(VALUE == 300) %>%
  summarise(n = n()) %>%
  pull(n)

cat("N observations of Diastolic BP equal 23 These observations are removed")

task_3a %>%
  filter(DESCRIPTION == "Diastolic Blood Pressure") %>%
  filter(VALUE == 23) %>%
  summarise(n = n()) %>%
  pull(n)

task_3a <- task_3a %>%
  filter(!((DESCRIPTION == "Diastolic Blood Pressure" & VALUE == 23 )|
           (DESCRIPTION == "Systolic Blood Pressure" & VALUE == 300))) 

task_3a %>%
  ggplot(aes(VALUE)) +
  geom_histogram(bins = 300) +
  facet_wrap(~DESCRIPTION)



ggplot(task_3a, aes(x = VALUE, color = DESCRIPTION)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density Plot of Blood Pressure", x = "Blood Pressure (mmHg)", y = "Density") 



task_3a_agg <- task_3a %>%
  group_by(ENCOUNTER, DESCRIPTION) %>%
  summarise(Average_BP = mean(VALUE, na.rm = TRUE), .groups = "drop")  # Using mean, but median also works

# Convert to wide format
task_3a_agg_wide <- task_3a_agg %>%
  tidyr::pivot_wider(names_from = DESCRIPTION, values_from = Average_BP)

# Rename columns for clarity
colnames(task_3a_agg_wide) <- c("encounter_ID", "Diastolic_BP", "Systolic_BP")

ggplot(task_3a_agg_wide, aes(x = Systolic_BP, y = Diastolic_BP)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(title = "Scatter Plot: Systolic vs. Diastolic BP per encounter", 
       x = "Systolic BP (mmHg)", 
       y = "Diastolic BP (mmHg)") 

task_3a %>%
  group_by(DESCRIPTION) %>%
  summarise(Mean = mean(VALUE),
            SD = sd(VALUE),
            Median = median(VALUE),
            IQR = IQR(VALUE),
            Min = min(VALUE),
            Max = max(VALUE)) %>%
  kbl(booktabs = T,
      caption = "Summary statistics of blood pressure values across the whole datset.")
```


# Task 3b

Using the cleaned data from the first task, explore and compare the distribution of BMI in patients with diagnosed hypertension

*   Assuming that patients never "loose" the diagnosis of hypertension 
*   Showing N patients with hypertension
*   Annotating observations with hypertension status, ensuring date is accounted for 
*   Calculating BMI values for those who do not have these recorded (younger patients). BMI was calculated by hand, using height and weight from the same encounter. encounters without height are removed, as imputation of height would be troublesome and inaccurate 
*   Showing histograms of BMI for patients with BMI and when BMI was calculated by hand
*   Showing BMI by hypertension status, and overlaid density plots
*   Showing summary metrics of BMI among patients with and without hypertension
*   Showing results of simple logistic regression with BMI as predictor and hypertension diagnosis as outcome (not really describing the distribution but additional way to describe the association)

```{r}
cat("N patients with hypertension")
conditions_clean %>%
  filter(str_detect(DESCRIPTION, "(?i)Hypertension")) %>%
  distinct(PATIENT, .keep_all = T) %>%
  pull(DESCRIPTION) %>% table()

# Pulling patients with hypertension and their first diagnosis 
hypertension_ids_dates <- conditions_clean %>%
  filter(str_detect(DESCRIPTION, "Hypertension")) %>%
  distinct(PATIENT, START) %>%
  group_by(PATIENT) %>%
  mutate(min_start = min(START)) %>%
  ungroup() %>%
  filter(START == min_start) %>%
  select(-min_start) %>%
  mutate(hypertension = T)

# Addding hypertension diagnosis, ensuring the timing of the diagnosis is taken into account
observations_clean <- observations_clean %>%
  left_join(hypertension_ids_dates, by = c("PATIENT")) %>%
  mutate(START = ifelse(is.na(START),
                        ("2030-01-01"),
                        as.character(START)),
         START = ymd(START)) %>%
  mutate(hypertension = ifelse(START <= DATE,
                               T, 
                               F)) %>%
  select(-START)

patients_with_BMI_values <- observations_clean %>%
  filter(DESCRIPTION == "Body Mass Index") %>%
  pull(PATIENT) %>% 
  unique()

cat("N patients without a BMI record. BMI is calculated for \n those patients using weight and height measurements")
((observations_clean$PATIENT %>% unique()) %nin% patients_with_BMI_values ) %>%
  sum()



cat("Body height is sometimes missing for encounters of patients without BMI. \n These encoutners are removed as there is enough data already for this comparison. \n Height could be imputed as mean, but for those young patients this \n may not work well. Carrying forward last value also may not work well. \n Therefore entries with weight but no height are removed")
BMI_by_hand <- observations_clean %>%
  filter(PATIENT %nin% patients_with_BMI_values) %>%
  filter(DESCRIPTION %in% c("Body Height", "Body Weight")) %>%
  tidyr::pivot_wider(id_cols = c(PATIENT, ENCOUNTER, DATE, hypertension), names_from = DESCRIPTION, values_from = VALUE) %>%
  filter(!is.na(`Body Height`)) %>%
  mutate(across(contains("Body"), ~ as.numeric(.)),
         BMI_by_hand = `Body Weight`/(`Body Height`/100)^2,
         VALUE = BMI_by_hand,
         DESCRIPTION = "Body Mass Index",
         by_hand = T) %>%
  select(DATE, PATIENT, ENCOUNTER, VALUE, DESCRIPTION, by_hand, hypertension)

task_3b <- observations_clean %>%
  filter(DESCRIPTION == "Body Mass Index") %>%
  select(DATE, PATIENT, ENCOUNTER, VALUE, DESCRIPTION, hypertension) %>%
  mutate(by_hand = F) %>%
  rbind(BMI_by_hand) %>%
  mutate(VALUE = as.numeric(VALUE))


task_3b %>%
  ggplot(aes(VALUE)) +
  geom_histogram(bins = 300) +
  labs(title = "Histogram of BMI in the whole sample")

BMI_by_hand %>%
  ggplot(aes(VALUE)) +
  geom_histogram(bins = 300) +
  labs(title = "Histogram of BMI in the subset where BMI was calculated by hand\nThis population is quite different from the rest.")

cat("Mean age across patients for whom BMI was missing vs not missing.\n This explains the small height and odd BMI distribution")
task_3b %>%
  left_join(patients_clean %>%
              select(Id, DATE_BIRTHDATE), by = c("PATIENT" = "Id")) %>%
  mutate(age = as.numeric(difftime(DATE, DATE_BIRTHDATE, units = "days"))/365.25) %>%
  group_by(by_hand) %>%
  summarise(mean_age = mean(age))


task_3b %>%
  ggplot(aes(VALUE)) +
  geom_histogram(bins = 300) +
  facet_wrap(~hypertension) +
  labs(title = "Histogram of BMI observations by hypertension status")


ggplot(task_3b, aes(x = VALUE, color = hypertension)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density Plot of BMI", x = "BMI", y = "Density") 



task_3b %>%
  group_by(hypertension) %>%
  summarise(Mean = mean(VALUE),
            SD = sd(VALUE),
            Median = median(VALUE),
            IQR = IQR(VALUE),
            Min = min(VALUE),
            Max = max(VALUE)) %>%
  kbl(booktabs = T,
      caption = "Summary statistics of BMI values across the hypertension status.")

model <- glm(hypertension ~ VALUE, data = task_3b, family = binomial)

cat("Results of logistic regression with hypertension as outcome and BMI as predictor variable")
tidy(model, exponentiate = TRUE, conf.int = TRUE)
```


# Task 4

Report the crude, and adjusted (to the UK population as much as possible) prevalence of
hypertension.


*   Not accounting for recovery from hypertension 
*   Including all patients, regardless if died within the period
*   Using UK estimates for mid 2022

```{r}
crude_prevalence <- conditions_clean %>%
  filter(str_detect(DESCRIPTION, "Hypertension")) %>%
  distinct(PATIENT) %>%
  nrow() / n_patients 

cat(paste0("Crude prevalence in the whole dataset (N = ", n_patients, ") : ", round(crude_prevalence*100, 2), "%"))

patients_hypertension <- conditions_clean %>%
  filter(str_detect(DESCRIPTION, "Hypertension")) %>%
  pull(PATIENT) %>%
  unique()


US_ests <- patients_clean %>%
  mutate(hypertension = Id %in% patients_hypertension) %>%
  group_by(age_group, GENDER) %>%
  mutate(n = n()) %>%
  group_by(hypertension, GENDER, age_group) %>%
  mutate(n_hypertensive = n()) %>%
  distinct(hypertension, n, n_hypertensive, age_group) %>%
  arrange(age_group, GENDER) %>%
  filter(hypertension) %>%
  mutate(hypertension_prevalence = n_hypertensive/n,
         GENDER = ifelse(GENDER == "F",
                                "Females",
                                "Males"))



UK_ests_today <- pop_estimates %>%
  filter(Sex %in% c("Males", "Females"),
         Age != "All Ages") %>%
  mutate(Age = str_replace(Age, "\\+", ""),
         Age = as.numeric(Age),
         age_group = factor(cut(Age, 
                                breaks = c(seq(0, 90, by = 10), Inf), 
                                labels = c("0-9", "10-19", "20-29", "30-39", "40-49", 
                                           "50-59", "60-69", "70-79", "80-89", "90 or more"),
                                right = FALSE), # Ensures inclusive lower bounds
                            ordered = TRUE)) %>%
  group_by(Sex, age_group) %>%
  summarise(sum_n = sum(`Mid-2022`)) 

UK_pop_total <- UK_ests_today$sum_n %>%
  sum()

adjusted_prevalence <- UK_ests_today %>%
  mutate(prop = sum_n / UK_pop_total,
         ) %>%
  inner_join(US_ests, by = c("Sex" = "GENDER",
                             "age_group")) %>%
  mutate(prev_times_weight = hypertension_prevalence * prop) %>%
  pull(prev_times_weight) %>%
  sum()

cat(paste0("Adjusted prevalence, to the UK population in mid-2022 is: ", round(adjusted_prevalence, 4)*100, "%"))
```
